---
title: "Homework 2"
author: "Melike Aksoy"
uni: "mma2277"
date: 2023-10-04
output: github_document
---

```{r, message=FALSE}
library(tidyverse)
library(readxl)
```


## Problem 1

We clean the 538 `pols` data, which provides information on the number
of national politicians who are democratic or republican at any given
time. There are some values for which `prez_gop` is `2` – these are
months in which Ford became President following Nixon’s resignation. In
the new `president` variable created as part of our data cleaning, we
code these as `gop` (same as values when `prez_gop` is `1`).

``` r
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez")) 
```

We also clean the 538 `snp` data, which contains information related to
Standard & Poor’s stock market index.

``` r
snp = 
  read_csv(
    "./data/fivethirtyeight_datasets/snp.csv",
    col_types = cols(date = col_date(format = "%m/%d/%y"))) |>
  separate(date, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    year = if_else(year > 2023, year - 100, year)) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, close) 
```

Finally, we tidy the `unemployment` data so that it can be merged with
the `pols` and `snp` datasets.

``` r
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```

Now we merge the three datasets!

``` r
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)
## tibble [822 × 13] (S3: tbl_df/tbl/data.frame)
##  $ year        : num [1:822] 1947 1947 1947 1947 1947 ...
##  $ month       : chr [1:822] "January" "February" "March" "April" ...
##  $ month_num   : int [1:822] 1 2 3 4 5 6 7 8 9 10 ...
##  $ gov_gop     : num [1:822] 23 23 23 23 23 23 23 23 23 23 ...
##  $ sen_gop     : num [1:822] 51 51 51 51 51 51 51 51 51 51 ...
##  $ rep_gop     : num [1:822] 253 253 253 253 253 253 253 253 253 253 ...
##  $ gov_dem     : num [1:822] 23 23 23 23 23 23 23 23 23 23 ...
##  $ sen_dem     : num [1:822] 45 45 45 45 45 45 45 45 45 45 ...
##  $ rep_dem     : num [1:822] 198 198 198 198 198 198 198 198 198 198 ...
##  $ president   : chr [1:822] "dem" "dem" "dem" "dem" ...
##  $ month_abb   : chr [1:822] "Jan" "Feb" "Mar" "Apr" ...
##  $ close       : num [1:822] NA NA NA NA NA NA NA NA NA NA ...
##  $ unemployment: num [1:822] NA NA NA NA NA NA NA NA NA NA ...
```

Notice that there are some `NA` values in the `close` and `unemployment`
variables, which indicate that the value of these variables is missing
at those locations.

Let’s talk about the 538 datasets. The `pols` data has 822 observations
and 11 variables and tells us about the party affiliation distribution
(democrat or republican) for governors and senators for a given year
from years 1947 to 2015. It also tells us whether the sitting president
was a democrat or republican. The `snp` data has 787 observations and 3
variables, ranging from years 1950 to 2015. The `unemployment` data has
816 observations and 3 variables ranging from years 1948 to 2015. In
Januarys in or after 1975 in which a democrat was president, the
**average unemployment rate was 6.57**. The average unemployment rate
over the same time period in which a republican was president was 6.47.


#########################################################################################

## Problem 2


#### Importing, Cleaning, and Organize the Mr.Trash Wheel Data


```{r, message=FALSE}
mrtrashwheel_data = read_excel("/Users/muazzezaksoy/Desktop/p8105_hw2_mma2277/data/202309 Trash Wheel Collection Data.xlsx", 
  sheet = 1, ##### Selecting the second excel sheet that we will use 
  skip = 1,  ##### Skipping the first row because it contains figures
  ) |>
  janitor::clean_names() |>
  drop_na() |>  ##### Omitting rows that do not include dumpster-specific data using drop_na                  function
  mutate(Trash_Wheel = "Mr") |> ##### creating a new variable for identification
  select(- year)|>  ##### Removing the year column because it year column include non-data    
  select(- x15, - x16) |> #### Removing the columns x15 and x16
  select(Trash_Wheel, everything()) #### Reorganizing columns to join datasets later                                            (putting  identifier variable as first column)
   
```


#### Creating homes_powered variable
mrtrashwheel_data = mrtrashwheel_data |> 
  mutate(homes_powered = (weight_tons*500)/30)


#### Importing, Cleaning, and Organize Professor Trash Wheel data

```{r, message=FALSE}
proftrashwheel_data = read_excel("/Users/muazzezaksoy/Desktop/p8105_hw2_mma2277/data/202309 Trash Wheel Collection Data.xlsx", 
  sheet = 2,    ##### Selecting the second excel sheet that we will use 
  skip = 1,     ##### Skipping the first row because it contains figures
  ) |>
  janitor::clean_names() |>
  drop_na() |>  ##### Omitting rows that do not include dumpster-specific data using drop_na                  function
  mutate(Trash_Wheel = "Professor") |> ##### creating a new variable for identification
  select(Trash_Wheel, everything()) |> ##### Reorganizing columns to join datasets later
  select(- year)  ##### Removing the year column because it year column include non-data                         entries(columns with notes)
```


#### Importing, Cleaning, and Organize Professor Trash Wheel data


```{r, message=FALSE}
Gwynnda_data = read_excel("/Users/muazzezaksoy/Desktop/p8105_hw2_mma2277/data/202309 Trash Wheel Collection Data.xlsx", 
  sheet = 4, ##### Selecting the second excel sheet that we will use 
  skip = 1, ##### Skipping the first row because it contains figures
  )|>
  janitor::clean_names() |>
  drop_na()|> ##### Omitting rows that do not include dumpster-specific data using drop_na                  function  
  select(- year) |> ##### Removing the year column
  mutate(Trash_Wheel = "Gwynnda") |> ##### Creating a new variable for identification
  select(Trash_Wheel, everything())  #### Reorganizing columns to join datasets later

```


#### Joining datasets( Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda)

```{r}
tidy_trashwheel_data = 
  bind_rows(mrtrashwheel_data, proftrashwheel_data, Gwynnda_data)

```

### Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in July of 2021?



# Problem 3

#### Importing, Cleaning, and Organizing MCI baseline data

```{r}
MCI_baseline_data = read_csv(file = "/Users/muazzezaksoy/Desktop/p8105_hw2_mma2277/data/data_mci/MCI_baseline.csv",
  skip=1) |>
  janitor::clean_names() |>
  mutate(sex = ifelse(sex == 1, "Male", "Female"),
        apoe4 = ifelse(apoe4 == 1, "Carrier", "Non-carrier") ) 
  
  tibble(MCI_baseline_data)
```


##### Importing, Cleaning, and Organizing MCI baseline data
```{r}
MCI_amyloid_data = read_csv(file = "/Users/muazzezaksoy/Desktop/p8105_hw2_mma2277/data/data_mci/MCI_amyloid.csv",
  skip=1) |>
  janitor::clean_names() |>
  rename(id=study_id)
```


#### Joining datasets (MCI_baseline_data, MCI_amyloid_data)

```{r}
MCI_combined= 
  inner_join(MCI_baseline_data, MCI_amyloid_data, by = "id")
```

#### Converting from WIDE TO LONG FORMAT

```{r}
MCI_combined_long = MCI_combined |> 
  pivot_longer(cols = starts_with("time_"),  # Specifying the columns to pivot
               names_to = "Times",            # Name of the new variable for column names
               values_to = "Biomarker")          # Name of the new variable for biomarker
```




