---
title: "Homework 3"
author: "Melike Aksoy"
uni: "mma2277"
date: 2023-10-14
output: github_document
---

### Loading the libraries including library for p8105.datasets

```{r, message=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1: 
The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illstrative examples of observations. Then, do or answer the following (commenting on the results of each)

### Load the data
```{r, message=FALSE}
library(p8105.datasets)
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```



Q1:
How many aisles are there, and which aisles are the most items ordered from?
```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

Explanations:
This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(user_id, order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.


Q2:
Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.
```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Q3:
Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.
```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```


Q4:
Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table)
```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```


##############################################################################################

## Problem 2

### Loading BRFSS data from p8105.datasets

```{r, message=FALSE}
data("brfss_smart2010") 
```


### 1) Data Cleaning
##### Format the data to use appropriate variable names
##### Cleaning names (to use appropriate variable names)
##### Focus on the “Overall Health” topic
##### Including only responses from Excellent to Poor

```{r, message=FALSE}
brfss_data=
  brfss_smart2010 |>
  janitor::clean_names()|>
  filter(topic=="Overall Health", response== "Excellent" | response == "Very good"| response =="Good" | response =="Fair" | response =="Poor") |>
  mutate(response=fct_relevel(response,c("Excellent", "Very good", "Good", "Fair", "Poor"))) |>
  rename(state=locationabbr , location=locationdesc)

```


### 2) Using this dataset, do or answer the following (commenting on the results of each):

#### Q1: 
In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r, message=FALSE}
brfss_data_2002 <- brfss_data |> 
  group_by(state, year) |>
  summarize(n_locations = n()) |> 
  filter(n_locations >= 7 & year== 2002)

brfss_data_2010 <- brfss_data |> 
  group_by(state, year) |>
  summarize(n_locations = n()) |> 
  filter(n_locations >= 7 & year== 2010)

brfss_state_2002=
  select(brfss_data_2002, state)

brfss_state_2010=
  select(brfss_data_2010, state)

```

This dataset contains `r nrow(brfss_data_2002)` states that were observed in 7 or more locations in 2002. For 2010, there are `r nrow(brfss_data_2010)` states that observed in 7 or more location. 


#### Q2: 
Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r, message=FALSE}
excellent_data <- brfss_data |>
  filter(response=="Excellent")|>
  group_by(year, state)|>
  summarize(average_data_value = mean(data_value, na.rm = TRUE))

ggplot(excellent_data, aes(x=year, y=average_data_value, color=state, group=state))+
  geom_line()+
    labs(
    title = "Average value over time within a state from 2002 to 2010",
    x = "Years",
    y = "Average data value",
    color = "States",
       )+
theme(legend.position = "right")
```

Explanations:
Average value over time within a state from 2002 to 2010 decreased for all the states. Groups of lines move together and share similar patterns. Most of the data clustured aroung average data value of 20 to almost 30. Between 2004 to 2006, there is decline in average data values in most of the states. Low WY seems to be an outlier. Overall, states that are yellow seems (WY,WV, WI, WA) to be having low average data value over time compared to other states. 


#### Q3: 
Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

```{r, message=FALSE}
brfss_data_pan <- brfss_data |>
  filter(state=="NY" & year %in% c(2006,2010)) 

 
  ggplot(brfss_data_pan, aes(x = data_value, fill = response)) +
  geom_density(alpha = .5) + 
  facet_grid(~ year) + 
  viridis::scale_fill_viridis(discrete = TRUE)+
  labs(title = "Distribution of data_value for responses (“Poor” to “Excellent”)",
       x="Data Value",
       y="Density")
```

Explanations:

In 2006, among the locations in NY state, the distribution of data value for responses shows that overall health that was reported were really different for each responses. The density plot for 2006 is right skewed which shows that data values are mostly concentrated on the left. The spread of responses excellent, very good, good are wide which shows high variability, while the spread of poor and fair responses are narrow which mean there is low variability in data values for those responses. 

For 2010,among the locations in NY state, the the distribution of data value for responses shows that overall health that was reported were more similar for each responses. The density plot for 2010 is right skewed which shows that data values are mostly concentrated on the left. The spread of good, very good, excellent, and fair responses are wide which shows high varibility. The category of poor response shows low variability. 

When we compare density plot of 2006 to 2010, we can see that data values for the responses become more wider which means that from 2006 to 2010 the variability increased between the responses. 




########################################################################################


## Problem 3

Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

####Importing Nhanes Demographic data

```{r, message=FALSE}
nhanes_dem = 
  read_csv(
    "/Users/muazzezaksoy/Desktop/p8105_hw3_mma2277/data/nhanes_covar (1).csv", 
    skip = 4, na = ".") |> 
  janitor::clean_names() |>
  drop_na(age, sex, education) |>
  filter(age >= 21) 

```


####Importing Nhanes Accelerometer data
```{r, message=FALSE}
nhanes_accl =
  read_csv(
    "/Users/muazzezaksoy/Desktop/p8105_hw3_mma2277/data/nhanes_accel (1).csv") |>
  janitor::clean_names()
```

####Combining two datasets
```{r, message=FALSE}
nhanes_comb=
  left_join(nhanes_dem,nhanes_accl,by="seqn")
```


#### Producing Reader-friendly table 
Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.

```{r, message=FALSE}
nhanes_comb |>
  select(education,sex) |>
  mutate(education = case_when(
    education == 1 ~ "Less than high school",
    education == 2 ~ "High school equivalent",
    education == 3 ~ "More than high school", 
    TRUE ~ as.factor(education)  
    )) |> 
  mutate(sex=case_match(sex,
     1 ~ "Male",
     2 ~ "Female")
     ) |>
  group_by(education, sex) |>
  summarize(n_obs = n())|> 
  knitr::kable(digits = 3)

```

Explanations:

From 250 participants in the NHANES study, only 231 were eligible for analysis due to missing values. The education level with highest number of males and females is more than high school. More than 1/3 of the participants attended more than high school. The lowest number of people were in the group with less than high school. In less than high school group, females and males have almost equal numbers. In high school equivalent group, females have the lowest number while males had high numbers. 

#### Visualization
```{r}
nhanes_comb$sex <- factor(nhanes_comb$sex, levels = c(1, 2), labels = c("Male", "Female"))
nhanes_comb$education <- factor(nhanes_comb$education, levels = c(1, 2, 3), labels = c("Less than high school", "High school equivalent" , "More than high school"))

ggplot(nhanes_comb, aes(x=age , y=education, fill=education))+
  geom_density_ridges(scale=.85)+
  facet_grid(~ sex)
  labs(title = "Age distribution by education and gender",
    x = "Age",
    fill = "Sex",)

```

Explanations:

Less than high school: The distribution for males are bimodal while distribution for females are unimodal. Females are more old than males for less than high school group. While males peaked in both between 25-35 and around 60-75, females peaked only around 60-75. This suggest that attending less than high school is more common in most of the age groups in males while for females only older ages have less than high school education.

High school equivalent, the distribution for males are wider than females. While males have bimodal distribution, females have unimodal distribution. The ages varied more for males than females. High school equivalent group for females mostly include people who are older than 50 while for males, the ages is almost equally distributed.

More than high school: For both males and females, the distribution is right skewed and almost unimodal. Distribution for males and females are similar which suggest that more than high school education is more common for both females and males. Also, peoplw who attend more than high school is younger. Highest point for the males and females are between 25 and around 35. 


#### Plotting total activities against age
Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r, message=FALSE}
nhanes_comb <- nhanes_comb %>%
  mutate(nhanes_comb, total_activity = rowSums(select(nhanes_comb,starts_with("min")))
           )
  relocate(nhanes_comb, seqn, sex, age, bmi, education, total_activity, everything())
```

```{r, message=FALSE}

nhanes_comb$sex <- factor(nhanes_comb$sex, levels = c(1, 2), labels = c("Male", "Female"))
nhanes_comb$education <- factor(nhanes_comb$education, levels = c(1, 2, 3), labels = c("Less than high school", "High school equivalent" , "More than high school"))

ggplot(nhanes_comb, aes(x = age, y = total_activity, color = sex, group=sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(~ education) +
  theme(legend.position = "right")+
  viridis::scale_fill_viridis(discrete = TRUE)+
  labs(title = "Total Activity of Different Age Groups",
       x="Age",
       y="Total Activity")

```


#### Creating new variable (24-hour activity) and plotting it to show 24 hour activity time courses for each education level
Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.


```{r, message=FALSE}
nhanes_comb <- nhanes_comb |>
mutate(nhanes_comb, hr_activity= total_activity/1440)|>
relocate( seqn, sex, age, bmi, education, total_activity, hr_activity, everything())


ggplot(nhanes_comb, aes(x = hr_activity, fill = sex)) +
  geom_density(alpha = .5) + 
  facet_grid(~education) + 
  viridis::scale_fill_viridis(discrete = TRUE) +
  theme(legend.position = "bottom")
```




