---
title: "Homework 3"
author: "Melike Aksoy"
uni: "mma2277"
date: 2023-10-14
output: github_document
---

### Loading the libraries including library for p8105.datasets

```{r, message=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1: 
The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illstrative examples of observations. Then, do or answer the following (commenting on the results of each)

### Load the data
```{r, message=FALSE}
library(p8105.datasets)
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```



Q1:
How many aisles are there, and which aisles are the most items ordered from?
```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

Explanations:
This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(user_id, order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.


Q2:
Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.
```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Q3:
Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.
```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```


Q4:
Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table)
```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```


##############################################################################################

## Problem 2

### Loading BRFSS data from p8105.datasets

```{r, message=FALSE}
data("brfss_smart2010") 
```


### 1) Data Cleaning
##### Format the data to use appropriate variable names
##### Cleaning names (to use appropriate variable names)
##### Focus on the “Overall Health” topic
##### Including only responses from Excellent to Poor

```{r, message=FALSE}
hw4_data=
  brfss_smart2010 |>
  janitor::clean_names()|>
  filter(topic=="Overall Health", response== "Excellent" | response == "Very good"| response =="Good" | response =="Fair" | response =="Poor") |>
  mutate(response=fct_relevel(response,c("Excellent", "Very good", "Good", "Fair", "Poor"))) |>
  rename(state=locationabbr , location=locationdesc)

```


### 2) Using this dataset, do or answer the following (commenting on the results of each):

### Q1: 
In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r, message=FALSE}
hw4_data |>
  group_by(state, year) |>
  summarize(n_locations = n()) |> 
  filter(n_locations >= 7 & year== 2002 | n_locations >= 7 & year== 2010)
```



### Q2: 
Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r, message=FALSE}
excellent_data <- hw4_data |>
  filter(response=="Excellent")|>
  group_by(year, state)|>
  summarize(average_data_value = mean(data_value, na.rm = TRUE))

ggplot(excellent_data, aes(x=year, y=average_data_value, color=state, group=state))+
  geom_point()+
  geom_line()+
    labs(
    title = "Average value over time within a state from 2002 to 2010",
    x = "Years",
    y = "Average data value",
    color = "States",
       )+
theme(legend.position = "right")
```

### Q3: 
Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State.

```{r, message=FALSE}
hw4_data_pan <- hw4_data |>
  filter(state=="NY" & year %in% c(2006,2010)) 

 
  ggplot(hw4_data_pan, aes(x = data_value, fill = response)) +
  geom_density(alpha = .5) + 
  facet_grid(~ year) + 
  viridis::scale_fill_viridis(discrete = TRUE)+
  labs(title = "Distribution of data_value for responses (“Poor” to “Excellent”)",
       x="Data Value",
       y="Density")
```

